---
title: "BarplotsINEMortalitySex"
output: html_document
date: '2022-12-27'
---

```{r setup, include=FALSE}
#--------------------- Import libraries
library(tidyr)
library(ggplot2)
library(reshape2)
library(dplyr)
```

## Import data

*Data Source*:

```{r cars}
data<-read.csv("Provincias_datos.csv", sep = ",")
# Glimpse of the dataset 
head(data)
```

## 1. Month Deaths Bar Chart

## Data preprocessing

Taking total cases, and month data, dropping missing values and translating month labels

```{r dp1}
datatotalmonth = data[which(data["Lugar"]=="Total"),]
datatotalmonth = datatotalmonth[which(data["Mes"]!="Total"),]
datatotalmonth = datatotalmonth[which(data["Sexo"]=="Total"),]
datatotalmonth = datatotalmonth %>% drop_na()
```

Subsetting and reformatting final dataframe

```{r sbs1}
datatotalmonth = datatotalmonth[,c("Mes", "Total")]

levels_order <-  c("January", "February", "March", "April", "May", "June", "July", "August", "September", "October", "November", "December")
datatotalmonth[,"Mes"] <- levels_order

# Necessary to reformat for ggplot
plotmonth = melt(datatotalmonth)
colnames(plotmonth) = c("Month", "Stratification", "Deaths")
```

## Plotting

```{r pressure, echo=FALSE}
#---------------------- Month plot


ggplot(data = data.frame(plotmonth), mapping = aes(x=Month,y=Deaths))+
  scale_x_discrete(limits = levels_order)+
  geom_bar(stat = "identity", position = 'dodge', color = "#7895B2", fill = "white", width = .3)+
  geom_path(aes(x = Month, y = Deaths), size = 1.5, color="red", group = 1, linewidth = .5)+
  ylab("Deaths")+
  ggtitle("Deaths per month in 2020")+
  theme(plot.title = element_text(size=16))+
  theme(axis.text.x = element_text(angle=45, hjust=1, size = 12),
        axis.text.y = element_text(size = 12),
        axis.title=element_text(size=14))
ggsave(filename = "deaths_month.tiff", width = 15, height = 10, device='tiff', dpi=700)
```

## 2. Month Sex Deaths Bar Chart

## Data preprocessing

Taking total cases, sex data and month data, dropping missing values and translating month and sex labels

```{r dp2}
datatotalmonthsex = data[which(data["Lugar"]=="Total"),]
datatotalmonthsex = datatotalmonthsex[which(data["Mes"]!="Total"),]
datatotalmonthsex = datatotalmonthsex %>% drop_na()
```

Subsetting and reformatting final dataframe

```{r sbs2}
datatotalmonthsex = datatotalmonthsex[,c("Mes", "Sexo","Total")]

# This time for translation of labels it is necessary to replicate the sex for each of the three sex labels per  instance
datatotalmonthsex["Sexo"] <- rep_len(c("Total", "Male", "Female"),3)

# This time for translation of labels it is necessary to replicate the months for each of the three sex labels per instance
levels_order <-  c(rep_len("January",3), rep_len("February",3), rep_len("March",3), rep_len("April",3), 
                   rep_len("May",3), rep_len("June",3), rep_len("July",3), rep_len("August",3), 
                   rep_len("September",3), rep_len("October",3), rep_len("November",3), rep_len("December",3))
datatotalmonthsex[,"Mes"] <- levels_order

# Necessary to reformat for ggplot
plotmonthsex = melt(datatotalmonthsex)
colnames(plotmonthsex) = c("Month", "Sex", "Stratification", "Deaths")
df=data.frame(plotmonthsex)
```

## Plotting

```{r 2 pressure, echo=FALSE}
#---------------------- Sex-Month plot


monthsex <- ggplot(data = df, mapping = aes(x=Month,y=Deaths))+
  scale_x_discrete(limits = levels_order)+
  geom_bar(data=df[df$Sex=="Total", ], stat = "identity", position = 'dodge', color = "#7895B2", fill = "white", width = 1.2)+
  geom_point(data=df[df$Sex!="Total", ],aes(color=Sex, x = as.factor(Month), y = Deaths, group=Sex), size = 2.5)+
  geom_path(data=df[df$Sex!="Total", ], aes(color=Sex, x = as.factor(Month), y = Deaths, group=Sex), size = 1.5, linewidth = .8)+
  ylab("Deaths")+
  ggtitle("Deaths per month in 2020")+
  theme(plot.title = element_text(size=16))+
  theme(axis.text.x = element_text(angle=45, hjust=1, size = 12),
        axis.text.y = element_text(size = 12),
        axis.title=element_text(size=14))
monthsex  + scale_fill_manual("legend", values = c("Female" = "#F96666", "Male" = "#7895B2"))

ggsave(filename = "deaths_pmonth_sex.tiff", width = 15, height = 10, device='tiff', dpi=700)
```

## 3. Distribution of Deaths per month boxplot

## Data preprocessing

Taking total cases, sex data and month data, dropping missing values and translating month and sex labels

```{r dp3}
datatotal = data[which(data["Lugar"]!="Total"),]
datatotal = datatotal[which(data["Mes"]!="Total"),]
datatotal = datatotal[which(data["Sexo"]=="Total"),]
datatotal = datatotal %>% drop_na()
```

Subsetting and reformatting final dataframe

```{r sbs3}
datatotal = datatotal[,c("Lugar", "Mes", "Total")]

# Gets number of provinces in the dataset for translation of month values through repetition
numprov <- length(unique(datatotal["Lugar"])[,1])

# This time for translation of labels it is necessary to replicate the months for each of the three sex labels per instance
levels_order <-  rep_len(c("January", "February", "March", "April", "May", "June", "July", "August", "September", "October", "November", "December"), numprov)
datatotal[,"Mes"] <- levels_order

# Necessary to reformat for ggplot
plottotalbp = melt(datatotal)
colnames(plottotalbp) = c("Province", "Month", "Stratification","Deaths")
df=data.frame(plottotalbp)
```

## Plotting

```{r 3 pressure, echo=FALSE}
#---------------------- Sex-Month plot


monthsex <- ggplot(data = df, mapping = aes(x=Month,y=Deaths))+
  scale_x_discrete(limits = c("January", "February", "March", "April", "May", "June", "July", "August", "September", "October", "November", "December"))+
  geom_boxplot()+
  ggtitle("Deaths per month in 2020")+
  theme(plot.title = element_text(size=16))+
  theme(axis.text.x = element_text(angle=45, hjust=1, size = 12),
        axis.text.y = element_text(size = 12),
        axis.title=element_text(size=14))
monthsex
ggsave(filename = "death_distribution_month.tiff", width = 15, height = 10, device='tiff', dpi=700)
```

## Plotting

```{r 4 pressure, echo=FALSE}
#---------------------- Sex-Month plot

monthsex <- ggplot(data = df, mapping = aes(x=Province,y=Deaths))+
  geom_boxplot()+
  ggtitle("Deaths per province in 2020")+
  theme(plot.title = element_text(size=16))+
  theme(axis.text.x = element_text(angle=45, hjust=1, size = 6),
        axis.text.y = element_text(size = 12),
        axis.title=element_text(size=14))
monthsex
ggsave(filename = "death_distribution_province.tiff", width = 15, height = 10, device='tiff', dpi=700)
```

## Grouping by Autonomic Community

```{r impccaa}

map_ccaa_prov <- read.csv("data/ccaa_provincia.csv")
colnames(map_ccaa_prov) <- c("CCAA","Lugar")
dataccaa <- merge(data,map_ccaa_prov,by="Lugar")

```

## Data preprocessing

Taking total cases, sex data and month data, dropping missing values and translating month and sex labels

```{r dp4}
dataccaa = dataccaa[which(data["Lugar"]!="Total"),]
dataccaa = dataccaa[which(data["Mes"]!="Total"),]
dataccaa = dataccaa[which(data["Sexo"]=="Total"),]
dataccaa = dataccaa %>% drop_na()
```

Subsetting and reformatting final dataframe

```{r sbs4}
dataccaa = dataccaa[,c("CCAA", "Mes", "Total")]

# Gets number of provinces in the dataset for translation of month values through repetition
numprov <- length(unique(datatotal["Lugar"])[,1])

# This time for translation of labels it is necessary to replicate the months for each of the three sex labels per instance
levels_order <-  rep_len(c("January", "February", "March", "April", "May", "June", "July", "August", "September", "October", "November", "December"), numprov)
datatotal[,"Mes"] <- levels_order

# Necessary to reformat for ggplot
plotccaabp = melt(dataccaa)
colnames(plotccaabp) = c("CCAA", "Month","Stratification","Deaths")
df=data.frame(plotccaabp)
```

## Plotting

```{r 5 pressure, echo=FALSE}
#---------------------- Sex-Month plot

monthccaa <- ggplot(data = df, mapping = aes(x=CCAA,y=Deaths))+
  geom_boxplot()+
  ggtitle("Deaths per CCAA in 2020")+
  theme(plot.title = element_text(size=16))+
  theme(axis.text.x = element_text(angle=45, hjust=1, size = 6),
        axis.text.y = element_text(size = 12),
        axis.title=element_text(size=14))
monthccaa
ggsave(filename = "death_distribution_ccaa.tiff", width = 15, height = 10, device='tiff', dpi=700)
```

```{r 6 pressure, echo=FALSE}
dataaccumulativeccaa = dataccaa %>% group_by(CCAA) %>%
  summarise(sum.Total = sum(Total))

# Necessary to reformat for ggplot
plotccaa_acc = melt(dataaccumulativeccaa)
colnames(plotccaa_acc) = c("CCAA", "Strat:Agg_Funct", "Deaths")
df=data.frame(plotccaa_acc)

#---------------------- Sex-Month plot

monthccaa <- ggplot(data = df, mapping = aes(x=CCAA,y=Deaths))+
  geom_bar(stat = "identity", position = 'dodge', color = "#7895B2", fill = "white", width = 0.7)+
    ggtitle("Deaths per CCAA in 2020")+
  theme(plot.title = element_text(size=16))+
  theme(axis.text.x = element_text(angle=45, hjust=1, size = 6),
        axis.text.y = element_text(size = 12),
        axis.title=element_text(size=14))
monthccaa
ggsave(filename = "death_sum_ccaa.tiff", width = 15, height = 10, device='tiff', dpi=700)
```
